---
title: "Active Learning and Learning Curve Extrapolation in celecx"
bibliography: references.bib
link-citations: true
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Active Learning and Learning Curve Extrapolation in celecx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
required_packages <- c("mlr3learners", "DiceKriging", "rpart", "lhs")
run_code <- all(vapply(required_packages, requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_code,
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  fig.align = "center"
)
```

# Introduction

`celecx` (Computer Experiment LEarning Curve eXtrapolation) provides infrastructure for batch-sequential computer experiments, that is, for settings where an expensive simulator is evaluated in successive batches and a regression model---the *surrogate* or *emulator*---is fit to the accumulating results [@SacksWelchMitchell1989; @SantnerWilliamsNotz2018].
Because every evaluation has a cost, two questions guide such an experiment:

1. *Which* configurations should be evaluated next, so that the surrogate improves as quickly as possible?
2. *How many* further evaluations will be needed before the surrogate reaches a desired level of quality?

The first question is the classical active-learning, or sequential design-of-experiments, question [@Settles2009].
The second question is usually answered informally, by looking at a plot of model error against sample size and extending the trend by eye.
`celecx` treats both questions as first-class problems.
It provides an active-learning toolkit that runs sequential designs and records their history, and it turns the resulting *learning curve*---surrogate quality as a function of the number of evaluated batches---into an ordinary supervised learning problem, so that competing extrapolation methods can be trained, compared, and used for principled "batches to target" forecasts.

This vignette takes both halves in turn. [Sections 2 to 4](#a-first-active-learning-example) introduce the active learning framework. [Section 2](#a-first-active-learning-example) begins with a basic example on a continuous domain. [Section 3](#active-learning-architecture) then examines in greater depth the different components underlying the active learning architecture, before [Section 4](#active-learning-on-a-finite-pool) turns to methods for pool-based active learning.
[Sections 5 to 8](#learning-curve-construction) then turn to learning curves, using a two-dimensional computer experiment as the main running example.
[Section 5](#learning-curve-construction) describes how to represent batchwise surrogate performance as an `mlr3` task, either by recording the learning curve online during the experiment or by reconstructing it offline through a replay of a finished archive.
Building on this representation, [Section 6](#learning-curve-extrapolation) introduces the functionality for extrapolating learning curves, while [Section 7](#extrapolator-evaluation) shows how different extrapolators can be compared through temporally ordered resampling.
Finally, [Section 8](#batches-to-target-forecasts-and-stopping-decisions) combines the resulting forecasts with a specified performance target to obtain a distribution over the batch in which the target will first be reached, thereby supporting stopping decisions.

## The `mlr3` Ecosystem

`celecx` builds on the `mlr3` ecosystem and reuses its vocabulary throughout.
The following summary should suffice to read this vignette; the packages' own documentation covers the details.

* **`paradox`** describes parameter spaces.
  A `ParamSet` is a collection of typed, possibly bounded parameters, constructed with `ps()` and the generators `p_dbl()`, `p_int()`, `p_fct()`, and `p_lgl()`.
  Parameter sets appear in two roles here: as the *domain* of a simulator (which inputs are admissible) and as its *codomain* (which outputs it produces, together with tags declaring what should be done with them).
* **`bbotk`** provides the black-box optimization backbone: an `Objective` couples an evaluation function with its domain and codomain, an `Archive` records all evaluated configurations together with their outcomes and batch numbers, and a `Terminator` decides when a run stops.
  `celecx` extends this toolkit towards pure learning: codomain targets may carry the tag `"learn"` instead of `"minimize"` or `"maximize"`, meaning that the output is observed and modelled but no best point is sought, and the `SearchInstance` class accepts such objectives.
* **`mlr3`** contributes the supervised-learning framework of tasks, learners, predictions, measures, and resampling schemes.
  It is used on two levels: regression learners (`LearnerRegr`) serve as surrogates during active learning, and `celecx` registers a dedicated task type `"lce"` under which learning curves themselves become tasks and extrapolation methods become learners.
* **`mlr3mbo`** supplies the surrogate and acquisition-function conventions from model-based optimization.
  A `SurrogateLearner` binds a regression learner to an archive, and an `AcqFunction` scores candidate points; `celecx` reuses both abstractions for active learning and registers its own acquisition functions.

Most objects in this ecosystem are `R6` classes with reference semantics: assigning an object to a second variable does not copy it, and modifications are visible through every reference.
Where an independent copy is needed, `$clone(deep = TRUE)` creates one; the `celecx` classes clone defensively wherever aliasing would be surprising.

## Setup

The examples use the Gaussian-process implementation from `DiceKriging` [@RoustantGinsbourgerDeville2012] via `mlr3learners`, regression trees from `rpart`, and Latin hypercube sampling from `lhs`.
Attaching `celecx` also attaches `paradox` and `bbotk`.

```{r setup, message = FALSE, warning = FALSE}
library("celecx")
library("mlr3")
library("mlr3learners")
library("data.table")
```

The optimization and learning packages report their progress through the `lgr` logging framework.
For a quiet document we raise the logging thresholds; interactive users may prefer to keep the progress output.

```{r logging}
lgr::get_logger("bbotk")$set_threshold("warn")
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("mlr3mbo")$set_threshold("warn")
```

# A First Active Learning Example

We begin with a one-dimensional "simulator", small enough that every aspect of a run can be plotted.
Throughout this section the expensive function is the test function of @GramacyLee2012, a damped oscillation superimposed on a quartic trend:

```{r simulator-1d}
simulator_1d <- function(x) {
  sin(10 * pi * x) / (2 * x) + (x - 1)^4
}
```

## Objective Definition

The simulator is wrapped as a `bbotk::Objective`.
The domain states that the single input `x` ranges over $[0.5, 2.5]$; the codomain states that the simulator returns one numeric output `y`, and the tag `"learn"` declares our intent: we want to *learn* the input--output relationship everywhere, not to locate an extremum.

```{r objective-1d}
objective_1d <- ObjectiveRFun$new(
  fun = function(xs) list(y = simulator_1d(xs$x)),
  domain = ps(x = p_dbl(lower = 0.5, upper = 2.5)),
  codomain = ps(y = p_dbl(tags = "learn"))
)

objective_1d$eval(list(x = 1.2))
```

`ObjectiveRFun` evaluates one configuration at a time; `ObjectiveRFunDt` is the batch-wise variant for functions that can process a whole `data.table` of configurations at once.
Codomains with `"minimize"` or `"maximize"` tags remain admissible---the same machinery can drive model-based optimization---but this vignette concentrates on the pure learning case.

## Running Active Learning

`optimize_active()` is the highest-level entry point.
Given an objective and an evaluation budget, it constructs an active-learning optimizer, runs it, and returns the objects needed for analysis.
As the surrogate we use Gaussian-process regression, the canonical emulator for smooth computer experiments; `nugget.stability` guards against numerical issues and `trace = FALSE` silences the underlying optimizer.

```{r first-run}
km_learner <- function() {
  lrn("regr.km", covtype = "matern5_2", nugget.stability = 1e-8,
    control = list(trace = FALSE))
}

set.seed(1L)

result_1d <- optimize_active(
  objective = objective_1d,
  n_evals = 24L,
  learner = km_learner(),
  se_method = "auto",
  batch_size = 4L,
  n_candidates = 200L
)
```

Behind this call, the following loop ran.
First, an initial design was evaluated; by default it has $4 \cdot d$ points for a $d$-dimensional search space, here four.
Then, until the budget of `n_evals = 24` evaluations was exhausted: the Gaussian process was refit on all evaluations so far, `n_candidates = 200` candidate points were sampled from the domain, each candidate was scored by the surrogate's predictive standard deviation, and the `batch_size = 4` highest-scoring candidates were evaluated as the next batch.
This is *uncertainty sampling*, the default strategy: it queries the simulator where the current model is least certain.

Since the Gaussian process predicts its own standard errors, `se_method = "auto"` used them directly.
The budget was expressed as a maximal number of evaluations for convenience; the `terminator` argument accepts any `bbotk::Terminator` (for example `trm("run_time")`) for richer stopping rules.

## Resulting Objects

The returned list carries the `SearchInstance`---the problem description together with the archive---and the optimizer.

```{r first-instance}
result_1d$instance
```

The archive is the central data structure of a run: one row per evaluated configuration, with the inputs, the observed output, and the batch in which the evaluation happened.

```{r first-archive}
head(result_1d$instance$archive$data[, .(x, y, batch_nr)], 8L)
```

The optimizer keeps its surrogate, so the final model is immediately available as an emulator.
Its `$predict()` method returns predictive means and standard errors for arbitrary input tables.

```{r first-emulator}
emulator <- result_1d$optimizer$surrogates$model
grid_1d <- data.table(x = seq(0.5, 2.5, length.out = 400L))
prediction_1d <- emulator$predict(grid_1d)
str(prediction_1d)
```

Figure 1 summarizes the run: the true function, the emulator with its 95% credible band, and the evaluated points shaded by batch number.
The batches alternate between the two difficult parts of the function: the oscillatory left, whose fine structure the model cannot yet interpolate, and the steep right flank. Each refit moves the widest part of the uncertainty band from one region to the other.
Within a batch the four points cluster tightly, a consequence of greedy batch construction that [Section 3.3](#batch-construction) returns to.

```{r first-plot, fig.height = 4, fig.cap = "Figure 1: Active learning on the one-dimensional test function. Points show evaluated configurations, shaded by batch; the shaded region is the emulator's 95% credible band after the final batch."}
archive_1d <- result_1d$instance$archive$data
batch_colors <- hcl.colors(max(archive_1d$batch_nr), "Viridis", rev = TRUE)

old_par <- par(mar = c(5.1, 4.1, 4.1, 4.6))
plot(grid_1d$x, simulator_1d(grid_1d$x), type = "n",
  xlab = "x", ylab = "y", main = "Uncertainty sampling with a GP surrogate")
polygon(c(grid_1d$x, rev(grid_1d$x)),
  c(prediction_1d$mean + 1.96 * prediction_1d$se,
    rev(prediction_1d$mean - 1.96 * prediction_1d$se)),
  col = adjustcolor("steelblue", alpha.f = 0.25), border = NA)
lines(grid_1d$x, simulator_1d(grid_1d$x), lwd = 2, col = "grey30")
lines(grid_1d$x, prediction_1d$mean, lwd = 2, col = "steelblue4")
points(archive_1d$x, archive_1d$y, pch = 19, cex = 1.1,
  col = batch_colors[archive_1d$batch_nr])
legend("topleft",
  legend = c("simulator", "emulator mean", "95% band"),
  col = c("grey30", "steelblue4", adjustcolor("steelblue", alpha.f = 0.25)),
  lwd = c(2, 2, 8), bty = "n")

# A compact color scale documents every batch without a separate legend entry.
plot_usr <- par("usr")
x_span <- diff(plot_usr[1:2])
y_span <- diff(plot_usr[3:4])
batch_x <- plot_usr[2] + c(0.04, 0.075) * x_span
batch_breaks <- seq(plot_usr[3] + 0.12 * y_span, plot_usr[4] - 0.12 * y_span,
  length.out = length(batch_colors) + 1L)
rect(batch_x[1L], head(batch_breaks, -1L), batch_x[2L], tail(batch_breaks, -1L),
  col = batch_colors, border = NA, xpd = NA)
rect(batch_x[1L], batch_breaks[1L], batch_x[2L], tail(batch_breaks, 1L),
  border = "grey30", xpd = NA)
text(mean(batch_x), tail(batch_breaks, 1L) + 0.035 * y_span, "batch",
  cex = 0.8, xpd = NA)
text(batch_x[2L] + 0.02 * x_span, batch_breaks[c(1L, length(batch_breaks))],
  labels = c(1L, length(batch_colors)), adj = c(0, 0.5), cex = 0.8, xpd = NA)
par(old_par)
```

# Active Learning Architecture

`optimize_active()` delegates the construction of its optimizer to the factory `optimizer_al()`, and every argument beyond the objective and the budget is in fact an argument of that factory.
Constructing the optimizer explicitly makes the structure of the method visible and allows the same configured strategy to be reused, inspected, or embedded in larger experiments.
The factory exposes three groups of choices: how the surrogate quantifies uncertainty, how candidates are generated, and how a batch is assembled from scored candidates.

```{r tree-optimizer}
tree_optimizer <- optimizer_al(
  learner = lrn("regr.rpart", minsplit = 2L, cp = 0),
  se_method = "bootstrap",
  n_bootstrap = 12L,
  batch_size = 2L,
  n_candidates = 200L,
  candidate_sampler = clx_sps("lhs"),
  multipoint_method = "greedy"
)
```

## Available Methods for Uncertainty Sampling

Uncertainty sampling needs predictive standard errors, but many attractive regression models do not provide them natively.
The `se_method` argument covers the three common situations:

* `"auto"` uses the learner's native `"se"` predict type when available, as with `regr.km` above, and falls back to `"bootstrap"` otherwise;
* `"bootstrap"` wraps the learner in `LearnerRegrBootstrapSE`, which trains an ensemble of `n_bootstrap` models on bootstrap resamples and reports the ensemble mean and standard deviation;
* `"quantile"` wraps a quantile-regression learner in `LearnerRegrQuantileSE`, which converts an inter-quantile range into a standard error.

Both wrappers are ordinary registered learners (`lrn("regr.bootstrap_se")`, `lrn("regr.quantile_se")`) and can be used outside active learning as well.
In the configuration above, a regression tree---a model with no notion of predictive variance---becomes usable for uncertainty sampling through a twelve-member bootstrap ensemble.

## Candidate Generation

In each proposal round the acquisition function scores a finite set of `n_candidates` candidate points, and the `candidate_sampler` determines how this set is drawn from the domain.
Samplers are `SpaceSampler` objects, collected in their own dictionary with the sugar constructors `clx_sps()` and `clx_spss()` (the `clx_` prefix marks `celecx` sugar, in analogy to `mlr3`'s `lrn()` and `msr()`):

```{r sampler-dict}
as.data.table(mlr_space_samplers)[, .(key, label)]
```

The configuration above replaces the default uniform sampling by a Latin hypercube design, matching the common practice of ranking space-filling candidates by model uncertainty.
Space-filling (`lhs`, `sobol`), geometric (`gsx`, the farthest-point strategy), and cluster-based (`kmeans`, `kmedoids`) samplers are interchangeable here, and the same objects also serve as initial-design generators.

## Batch Construction

When `batch_size` exceeds one, taking the top-scoring candidates ("greedy") risks proposing near-duplicates, since neighbouring points have similar acquisition scores.
The `multipoint_method` argument offers three sequential alternatives: `"local_penalization"` discounts scores near already-selected points [@GonzalezDaiHennig2016], `"diversity"` adds an explicit distance-based diversity term, and `"constant_liar"` inserts pseudo-observations at selected points and rescores, following @GinsbourgerLeRicheCarraro2010.
The tightly clustered same-batch points in Figure 1 show the phenomenon these methods address; for the small batches used in this vignette the greedy choice remains adequate, and we keep it.

## Optimizer Execution

A configured optimizer is passed to `optimize_active()` in place of the learner arguments.
The run below repeats the experiment of [Section 2](#a-first-active-learning-example) with the tree ensemble, a smaller batch size, and a larger budget; we will return to its archive in [Section 5](#learning-curve-construction).

```{r tree-run}
set.seed(2L)

result_tree <- optimize_active(
  objective = objective_1d,
  n_evals = 32L,
  optimizer = tree_optimizer
)

result_tree$instance
```

## Under the Hood: `OptimizerAL`

The factory itself only wires together components of the `OptimizerAL` class, and the same wiring can be written out explicitly.
An `OptimizerAL` owns named *surrogates* and *acquisition functions*, an *initial-design sampler*, and a *proposer* that turns the current state into the next batch; proposers refer to surrogates and acquisition functions by their registry names.
The following construction is, up to a technicality, what `optimizer_al()` assembled above (the factory additionally wraps the samplers so that the same optimizer serves both continuous domains and the finite pools of [Section 4](#active-learning-on-a-finite-pool)).

```{r manual-optimizer}
se_learner <- lrn("regr.bootstrap_se", learner = lrn("regr.rpart", minsplit = 2L, cp = 0))
se_learner$param_set$set_values(n_bootstrap = 12L)
se_learner$predict_type <- "se"

optimizer_manual <- OptimizerAL$new(
  proposer = ALProposerScore$new(
    acq_id = "sd",
    surrogate_id = "model",
    candidate_sampler = clx_sps("lhs"),
    n_candidates = 200L
  ),
  surrogates = list(
    model = mlr3mbo::SurrogateLearner$new(learner = se_learner, archive = NULL)
  ),
  acq_functions = list(sd = mlr3mbo::acqf("sd")),
  init_sampler = clx_sps("uniform"),
  result_assigner = ResultAssignerNull$new()
)
optimizer_manual$param_set$set_values(batch_size = 2L)

optimizer_manual
```

The print output shows how the components' configuration surfaces in the optimizer's parameter set under prefixes such as `proposer.` and `surrogate_model.`.
This decomposition is also the extension interface: a new active-learning method is typically a new `AcqFunction` (possibly building on the distance-aware `AcqFunctionDist` family and an `ALDistance` from the `mlr_al_distances` dictionary, which includes a Gower distance for mixed continuous--categorical spaces) or a new `ALProposer`, wired exactly as above.
The proposer classes shipped with the package---`ALProposerScore`, `ALProposerSequentialScore`, `ALProposerSequentialReference`, `ALProposerPseudoLabel`, and the round-robin `ALProposerPortfolio`---cover single-pass scoring, the sequential batch heuristics of [Section 3.3](#batch-construction), distance-based rescoring, pseudo-label updates, and method combination.

# Active Learning on a Finite Pool

So far, candidates could be placed anywhere in the domain.
In many applications the choice is instead restricted to a finite *pool* of configurations: a library of prepared meshes or material compositions, a factorial catalogue of admissible settings, or---particularly relevant for method development---a dataset of already-computed simulations on which an active-learning strategy is to be *replayed* as if the outcomes were still unknown.
Pool-based active learning is also where the best-known methods from the regression active-learning literature are formulated.

## Pool Objectives

`ObjectiveDataset` wraps a pre-evaluated table as an objective that evaluates by lookup and rejects configurations outside the table.
We discretize the simulator of [Section 2](#a-first-active-learning-example) into a pool of 101 candidate runs:

```{r pool-objective}
pool <- data.table(x = seq(0.5, 2.5, length.out = 101L))
pool[, y := simulator_1d(x)]

objective_pool <- ObjectiveDataset$new(
  dataset = pool,
  domain = ps(x = p_dbl(lower = 0.5, upper = 2.5)),
  codomain = ps(y = p_dbl(tags = "learn"))
)
```

Such objectives carry the property `"pool_restricted"`, and `OptimizerAL` reacts to it: initial designs and candidate sets are then drawn from the not-yet-evaluated pool rows rather than from the continuous domain, so the optimizers of the previous sections work unchanged.
(By default an evaluated pool row is never proposed again; the optimizer's `replace_samples` parameter relaxes this.)
For on-demand evaluation restricted to a candidate list, `ObjectivePoolRFun` and `ObjectivePoolWrapper` provide the same behaviour around a live function.

## Available Methods

`optimizer_pool_al()` constructs the standard pool-based methods, again as `OptimizerAL` wirings:

| `method` | Strategy | Reference | Learner used |
|---|---|---|---|
| `"random"` | uniform random sampling | -- | none |
| `"gsx"` | greedy sampling in input space | @WuLinHuang2019 | none |
| `"gsy"` | greedy sampling in output space | @WuLinHuang2019 | predictions |
| `"igs"` | improved greedy sampling, input $\times$ output | @WuLinHuang2019 | predictions |
| `"qbc"` | query by committee via bootstrap disagreement | @SeungOpperSompolinsky1992; @RayChaudhuriHamey1995 | committee |
| `"ideal"` | inverse-distance weighted residuals plus exploration | @Bemporad2023 | predictions |

GSx spreads evaluations by always taking the pool point farthest from the evaluated set; it needs no model at all.
GSy transfers the same greedy logic to the response scale, preferring candidates whose *predicted* output differs most from all observed outputs, and iGS combines both distances multiplicatively.
QBC queries where a bootstrap committee of models disagrees most, and IDEAL scores candidates by inverse-distance-weighted residuals together with a geometric exploration term whose weight is set by `delta`.
Each method comes with its customary initialization, which `init_method` can override (`"gsx"`, `"random"`, or `"kmeans"`, the latter selecting cluster-representative pool points).

## Comparison of Selection Behavior

We run four of the methods on the pool with identical budgets: sixteen evaluations, of which the first four form the initial design.
The model-based methods use the Gaussian process as their learner.

```{r pool-runs}
pool_optimizers <- list(
  random = optimizer_pool_al("random", n_init = 4L),
  gsx = optimizer_pool_al("gsx", n_init = 4L),
  igs = optimizer_pool_al("igs", learner = km_learner(), n_init = 4L),
  ideal = optimizer_pool_al("ideal", learner = km_learner(), n_init = 4L)
)

pool_results <- list()
for (method in names(pool_optimizers)) {
  set.seed(4L)
  pool_results[[method]] <- optimize_active(
    objective = objective_pool,
    n_evals = 16L,
    optimizer = pool_optimizers[[method]]
  )
}
```

Figure 2 shows which pool points each method selected.
Numbers and colours both encode archive order, from yellow (early) to purple (late): rows 1--4 are the jointly evaluated initial design, while rows 5--16 are selected sequentially because these runs use a batch size of one.
Random sampling scatters without regard to geometry, and GSx produces a near-equispaced space-filling sequence.
The model-based methods behave differently: iGS and IDEAL both shift effort towards the right half of the domain, where the steep trend produces large output differences and large residuals, respectively.

```{r pool-plot, fig.height = 6, fig.cap = "Figure 2: Selection behaviour of four pool-based methods on the same candidate pool. Colours encode archive order; a black outline marks the four points in each jointly evaluated initial design."}
plot_pool_run <- function(result, title, order_colors) {
  archive <- result$instance$archive$data
  plot(pool$x, pool$y, type = "l", col = "grey60",
    xlab = "x", ylab = "y", main = title)
  points(pool$x, pool$y, pch = 16, cex = 0.35, col = "grey75")
  points(archive$x, archive$y, pch = 19, cex = 1.1,
    col = order_colors[seq_len(nrow(archive))])
  initial <- archive$batch_nr == min(archive$batch_nr)
  points(archive$x[initial], archive$y[initial], pch = 21, cex = 1.2,
    col = "black", bg = order_colors[which(initial)], lwd = 1)
}

plot_order_scale <- function(order_colors) {
  n_evals <- length(order_colors)
  plot.new()
  plot.window(xlim = c(0, 1), ylim = c(0, n_evals), xaxs = "i", yaxs = "i")
  breaks <- seq(0.15 * n_evals, 0.85 * n_evals, length.out = n_evals + 1L)
  rect(0.05, head(breaks, -1L), 0.4, tail(breaks, -1L),
    col = order_colors, border = NA)
  rect(0.05, breaks[1L], 0.4, tail(breaks, 1L), border = "grey30")
  text(0.225, tail(breaks, 1L) + 0.6, "evaluation\nnumber", cex = 0.85)
  label_y <- c(mean(breaks[1:2]), mean(tail(breaks, 2L)))
  text(0.55, label_y, labels = c(1L, n_evals),
    adj = c(0, 0.5), cex = 0.95)
}

method_titles <- c(random = "Random", gsx = "GSx", igs = "iGS", ideal = "IDEAL")
n_pool_evals <- max(vapply(pool_results, function(result) {
  nrow(result$instance$archive$data)
}, integer(1)))
pool_order_colors <- hcl.colors(n_pool_evals, "Viridis", rev = TRUE)

old_par <- par(no.readonly = TRUE)
layout(matrix(c(1, 2, 5, 3, 4, 5), nrow = 2, byrow = TRUE),
  widths = c(1, 1, 0.19))
par(mar = c(4, 4, 2.5, 1), oma = c(0, 0, 3, 0))
for (method in names(pool_results)) {
  plot_pool_run(pool_results[[method]], method_titles[[method]], pool_order_colors)
}
par(mar = c(4, 0, 2.5, 0))
plot_order_scale(pool_order_colors)
mtext("Selection Behavior of Pool-Based Methods", side = 3, outer = TRUE,
  line = 1, font = 2, cex = 1.2)
par(old_par)
```

Which of these strategies actually learns fastest cannot be read off a selection plot; it is a question about the resulting model quality over time.
That question leads directly to the second half of the package.

# Learning Curve Construction

An active-learning run does not usually end because the model has become good enough; it ends because the budget is exhausted.
The economically relevant question during an active-learning run is prospective: given how quality has developed so far, how many further batches will be needed to reach the target?
The remainder of this vignette develops the `celecx` answer on a single running example.
The first step, taken in this section, is to make "quality so far" a concrete object: one surrogate-performance value per evaluated batch, recorded during or after the run.

## A Two-Dimensional Computer Experiment

The running example is a smooth response surface over $[0, 1]^2$, two broad overlapping rises on a mild trend, standing in for, say, a simulated process yield over two design parameters.
We aim for an emulator whose mean absolute error, over the whole design region, falls below a set target.

```{r simulator-2d}
simulator_2d <- function(x1, x2) {
  1000 * (exp(-2 * ((x1 - 0.35)^2 + (x2 - 0.6)^2)) +
    0.5 * exp(-3 * ((x1 - 0.85)^2 + (x2 - 0.15)^2)) +
    0.3 * x1)
}

objective_2d <- ObjectiveRFun$new(
  fun = function(xs) list(y = simulator_2d(xs$x1, xs$x2)),
  domain = ps(x1 = p_dbl(0, 1), x2 = p_dbl(0, 1)),
  codomain = ps(y = p_dbl(tags = "learn"))
)
```

Measuring surrogate quality requires a held-out regression task on which the surrogate is scored.
In a synthetic study we can afford a dense grid of true function values; in a real application the same role is played by whatever reference data are available, for example a reserved set of past simulator runs.
The held-out task is an ordinary `mlr3` regression task:

```{r test-task-2d}
test_grid <- CJ(x1 = seq(0, 1, length.out = 24L), x2 = seq(0, 1, length.out = 24L))
test_grid[, y := simulator_2d(x1, x2)]
test_task_2d <- as_task_regr(test_grid, target = "y", id = "held_out_2d")
```

## Online Curve Recording

`CallbackSurrogatePerformance` hooks into the run and, after every evaluated batch including the initial design, scores a named surrogate of the optimizer on the held-out task.
The callback is retrieved from the callback dictionary via `clbk()`; `surrogate_id = "model"` refers to the surrogate registry of `optimizer_al()`, and any list of regression measures can be tracked.

```{r experiment-run}
progress <- clbk("celecx.surrogate_performance",
  surrogate_id = "model",
  task = test_task_2d,
  measures = list(mae = msr("regr.mae"), rsq = msr("regr.rsq"))
)

set.seed(3L)

result_experiment <- optimize_active(
  objective = objective_2d,
  n_evals = 60L,
  learner = km_learner(),
  se_method = "auto",
  batch_size = 4L,
  n_candidates = 400L,
  callbacks = list(progress)
)
```

The run evaluated an initial design of eight points ($4 \cdot d$ with $d = 2$) followed by thirteen batches of four simulator runs each.
The callback accumulated one row per batch:

```{r experiment-progress}
progress$data[, .(batch_nr, n_evals, mae = signif(mae, 3), rsq = round(rsq, 4))]
```

The two measures tell complementary stories.
$R^2$ saturates almost immediately---it exceeds $0.999$ within a handful of batches---while the mean absolute error keeps falling by orders of magnitude long after that.
For emulator construction, where pointwise accuracy matters, the absolute error is the more informative quantity, and we use the MAE curve from here on.
Figure 3 summarizes the run.
The design ends up close to space-filling---for a surface this smooth, predictive uncertainty is governed mainly by data density---and the error decays over roughly three orders of magnitude, approximately linearly on the logarithmic scale with a gradually flattening slope.

```{r experiment-plot, fig.height = 4, fig.cap = "Figure 3: The two-dimensional computer experiment. Left: simulator contours with the evaluated design, shaded by batch. Right: held-out mean absolute error of the emulator after each batch."}
archive_experiment <- result_experiment$instance$archive$data
n_experiment_batches <- max(archive_experiment$batch_nr)
experiment_colors <- hcl.colors(n_experiment_batches, "Viridis", rev = TRUE)

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
x1_seq <- seq(0, 1, length.out = 60L)
x2_seq <- seq(0, 1, length.out = 60L)
contour(x1_seq, x2_seq, outer(x1_seq, x2_seq, simulator_2d),
  nlevels = 10, col = "grey60", xlab = "x1", ylab = "x2", main = "Design")
points(archive_experiment$x1, archive_experiment$x2, pch = 19, cex = 0.8,
  col = experiment_colors[archive_experiment$batch_nr])
plot(progress$data$batch_nr, progress$data$mae, log = "y", type = "b", pch = 19,
  xlab = "batch", ylab = "held-out MAE", main = "Learning curve")
par(old_par)
```

## Learning Curves as `mlr3` Tasks

The callback converts its records into a `TaskLCE`, the task type under which learning curves become `mlr3` data.
Two arguments matter here.
The `measure` argument selects which tracked measure becomes the prediction target.
The `link` argument declares the scale on which forecasting models should represent the curve and its uncertainty: a MAE is positive and decays multiplicatively, so we model it on the logarithmic scale, which also guarantees that predictive distributions respect the lower bound of zero.
The helper `lce_link_from_range()` suggests a link from a measure's theoretical range (`"log"` for losses on $(0, \infty)$, `"logit"` for scores on $(0, 1)$, `"identity"` otherwise):

```{r task-lce}
lce_link_from_range(msr("regr.mae")$range)

task_experiment <- progress$task(measure = "mae", link = "log", id = "experiment_2d")
task_experiment
```

The task has one row per *archive evaluation*, not per batch; all rows of a batch share that batch's performance value.
Its single feature is the batch number, which is all that a forecasting model needs at prediction time---forecasting the curve at future batches must not require knowing which points those batches will contain.

```{r task-lce-data}
head(task_experiment$data(), 3L)
```

The archive's input and output columns are nevertheless carried along, under the dedicated column roles `archive_x` and `archive_y` (accessible via `$archive_x_data()` and `$archive_y_data()`), because advanced extrapolators inspect the archive during training.
For the same reason the task stores run provenance: the search space and codomain of the originating run, the regression measure that produced the target (`task_experiment$measure`), the candidate pool for pool-based runs (`task_experiment$pool`, `NULL` here), and the link name (`task_experiment$link`).
[Section 6.3](#policy-aware-forward-simulation) shows a forecaster that consumes all of it.

## Offline Curve Reconstruction

Attaching the callback requires having thought of it before the run.
`replay_surrogate_performance()` removes that requirement: given a finished archive, it refits a surrogate on each batch prefix, scores it on the held-out task, and returns the same kind of `TaskLCE`.
Beyond rescuing untracked runs, replay decouples the curve from the run in two useful ways: the replayed surrogate need not be the one that drove the design, so several emulator choices can be scored on the same trace, and runs whose strategy maintained no model at all---GSx or random sampling, say---still yield learning curves.

The tree-ensemble run of [Section 3](#active-learning-architecture) was not tracked; we reconstruct its curve now, scoring against a dense held-out grid of the one-dimensional function.

```{r replay}
dense_1d <- data.table(x = seq(0.5, 2.5, length.out = 201L))
dense_1d[, y := simulator_1d(x)]
test_task_1d <- as_task_regr(dense_1d, target = "y", id = "held_out_1d")

tree_se_learner <- lrn("regr.bootstrap_se", learner = lrn("regr.rpart", minsplit = 2L, cp = 0))
tree_se_learner$param_set$set_values(n_bootstrap = 12L)
tree_se_learner$predict_type <- "se"

set.seed(21L)
task_tree <- replay_surrogate_performance(
  archive = result_tree$instance$archive,
  learner = tree_se_learner,
  task = test_task_1d,
  measures = list(mae = msr("regr.mae")),
  link = "log",
  id = "surrogate_1d"
)
```

```{r replay-plot, fig.height = 3.2, fig.cap = "Figure 4: Learning curve of the [Section 3](#active-learning-architecture) tree-ensemble run, reconstructed offline by replaying its archive."}
tree_curve <- unique(task_tree$data()[, .(batch_nr, mae)])
plot(tree_curve$batch_nr, tree_curve$mae, log = "y", type = "b", pch = 19,
  xlab = "batch", ylab = "held-out MAE", main = "Replayed learning curve (tree ensemble)")
```

This curve looks qualitatively different from the two-dimensional run's: after early gains it flattens out near a plateau, the bias floor of a piecewise-constant model on a smooth function.
We keep both tasks; their contrast becomes instructive in [Section 7](#extrapolator-evaluation).

# Learning Curve Extrapolation

A `TaskLCE` is a task, so extrapolation methods are learners: they train on the observed prefix of a curve and predict performance at future batch numbers.
`celecx` registers them under the `lce.` prefix in the standard learner dictionary:

```{r lce-zoo}
grep("^lce\\.", mlr_learners$keys(), value = TRUE)
```

The families are, in increasing order of structure:
featureless baselines (`lce.featureless` predicts the best, last, or average observed value forever);
a local trend baseline (`lce.rolling_slope` extends a straight line fit to the most recent batches);
parametric curve families from the learning-curve literature [@VieringLoog2023]---`lce.parametric_exponential` fits $f(b) = c + a e^{-kb}$, `lce.parametric_power_law` fits $f(b) = c + a b^{-k}$, and `lce.parametric_log` and `lce.parametric_logistic` fit logarithmic and sigmoidal shapes;
monotone nonparametric fits (`lce.isotonic`, and `lce.spline_monotone` if the `scam` package is installed), which impose the shape constraint without committing to a family;
wrappers that endow a base learner with distributional predictions (`lce.bootstrap`, `lce.conformal`);
and the forward-simulation learner `lce.simulate`, treated in [Section 6.3](#policy-aware-forward-simulation).
All of them fit the curve on the task's link scale, here the log scale.

## Mid-Run Forecasts

To use the run as an honest test bed, we place ourselves at batch 8---roughly its midpoint, with MAE just above $0.7$---and forecast from there, keeping the remaining batches for comparison.
Restricting a task to a batch prefix is a matter of filtering rows:

```{r prefix}
forecast_batch <- 8L
task_past <- task_experiment$clone(deep = TRUE)
task_past$filter(task_past$row_ids[task_past$batch_nrs <= forecast_batch])

realized <- unique(task_experiment$data()[, .(batch_nr, mae)])
```

We train four point forecasters on the prefix and predict the full range of batches.
Prediction at new batch numbers uses the standard `predict_newdata()` mechanism; note that `batch_nr` must be supplied as an integer column.

```{r point-forecasts}
point_learners <- list(
  featureless = lrn("lce.featureless"),
  rolling_slope = lrn("lce.rolling_slope"),
  exponential = lrn("lce.parametric_exponential"),
  power_law = lrn("lce.parametric_power_law")
)

batch_grid <- data.table(batch_nr = 1:16L)
point_forecasts <- sapply(point_learners, function(learner) {
  learner$train(task_past)
  learner$predict_newdata(batch_grid)$response
})
```

```{r forecast-plot, fig.height = 4.5, fig.cap = "Figure 5: Point forecasts from batch 8. Filled circles are the batches available to the forecasters, open circles the realized continuation."}
forecast_colors <- hcl.colors(5, "Dark 3")[1:4]
matplot(batch_grid$batch_nr, point_forecasts, log = "y", type = "l", lty = 1,
  lwd = 2, col = forecast_colors, ylim = range(realized$mae, point_forecasts),
  xlab = "batch", ylab = "held-out MAE",
  main = "Extrapolating the learning curve")
abline(v = forecast_batch + 0.5, lty = 3, col = "grey50")
points(realized[batch_nr <= forecast_batch], pch = 19)
points(realized[batch_nr > forecast_batch], pch = 1)
legend("bottomleft", legend = colnames(point_forecasts), col = forecast_colors,
  lwd = 2, bty = "n")
```

The four methods embody different beliefs about what comes next, and the figure makes the differences concrete.
The featureless baseline (predicting the best value seen so far) denies any further progress.
The rolling slope continues the recent log-linear trend indefinitely and therefore predicts the fastest progress.
The two parametric families interpolate between these extremes: both admit deceleration, but the exponential's fixed asymptote, estimated from an early prefix, is pessimistic about the later batches, while the power law's slower decay tracks the realized continuation more closely here.
None of this is specific to the example---which family extrapolates best depends on the curve, which is exactly why [Section 7](#extrapolator-evaluation) treats the choice as an empirical model-selection problem.

## Distributional Forecasts

Point forecasts are rarely enough: a stopping decision needs to know how *certain* the forecast is.
`LearnerLCE` objects therefore support distributional predict types beyond `"response"`:

* `"se"` adds two standard errors on the link scale: `se`, the total predictive standard deviation of a future *realized* curve value, and `se_epistemic`, the narrower uncertainty of the mean curve itself.
  The distinction matters because a realized batch value scatters around the mean curve with residual noise even if the curve were known exactly.
* `"quantiles"` returns predictive quantiles on the natural scale.
* `"samples"` returns joint sample paths, which preserve the dependence between future batches.
* `"target_reached"` returns, per batch, the probability that a configurable target has been reached.

The parametric learners derive `se` analytically from the curve fit.
A more assumption-averse route is the residual-bootstrap wrapper `lce.bootstrap`, which refits its base learner on many resampled versions of the curve and reads uncertainty off the spread of the refits; being sample-based, it supports all predict types including joint paths.
We wrap the power law:

```{r bootstrap-forecaster}
forecaster <- lrn("lce.bootstrap",
  learner = lrn("lce.parametric_power_law"),
  n_bootstrap = 100L, seed = 1L)
forecaster$predict_type <- "se"
forecaster$train(task_past)

prediction_se <- as.data.table(forecaster$predict_newdata(batch_grid))
tail(cbind(batch_grid, prediction_se[, .(response = signif(response, 3),
  se = signif(se, 3), se_epistemic = signif(se_epistemic, 3))]), 4L)
```

On the log link, a central 90% band around the predictive median is obtained as $\text{response} \cdot \exp(\pm z_{0.95} \cdot \text{se})$.
Figure 6 shows the epistemic and total bands together with a handful of joint sample paths, and the realized continuation for reference.

```{r fan-plot, fig.height = 4.5, fig.cap = "Figure 6: Distributional forecast of the bootstrap power law from batch 8: central 90% bands from the epistemic (dark) and total (light) standard errors, thin lines showing 15 joint sample paths of the realized future curve, and the realized continuation (open circles)."}
z90 <- qnorm(0.95)
forecaster$predict_type <- "samples"
future_grid <- data.table(batch_nr = (forecast_batch + 1L):16L)
sample_paths <- forecaster$predict_newdata(future_grid)$samples

plot(realized$batch_nr, realized$mae, log = "y", type = "n",
  xlim = c(1, 16), ylim = range(realized$mae) * c(0.2, 2),
  xlab = "batch", ylab = "held-out MAE", main = "Bootstrap power-law forecast")
polygon(c(batch_grid$batch_nr, rev(batch_grid$batch_nr)),
  c(prediction_se$response * exp(z90 * prediction_se$se),
    rev(prediction_se$response * exp(-z90 * prediction_se$se))),
  col = adjustcolor("darkorange", alpha.f = 0.15), border = NA)
polygon(c(batch_grid$batch_nr, rev(batch_grid$batch_nr)),
  c(prediction_se$response * exp(z90 * prediction_se$se_epistemic),
    rev(prediction_se$response * exp(-z90 * prediction_se$se_epistemic))),
  col = adjustcolor("darkorange", alpha.f = 0.35), border = NA)
matlines(future_grid$batch_nr, sample_paths[, 1:15],
  lty = 1, lwd = 0.5, col = adjustcolor("grey30", alpha.f = 0.35))
lines(batch_grid$batch_nr, prediction_se$response, lwd = 2, col = "darkorange3")
abline(v = forecast_batch + 0.5, lty = 3, col = "grey50")
points(realized[batch_nr <= forecast_batch], pch = 19)
points(realized[batch_nr > forecast_batch], pch = 1)
```

The alternative wrapper `lce.conformal` calibrates a constant-width band on a hold-out suffix of the training batches by split-conformal inference, trading the Gaussian working assumption for finite-sample marginal coverage at its configured level.

## Policy-Aware Forward Simulation

All learners so far look only at the shape of the observed curve.
`lce.simulate` uses the run provenance stored in the task instead: at training time it fits an *oracle* regression model on the archive's input--output pairs, and at prediction time it re-runs the configured active-learning optimizer forward from the training archive, using the oracle in place of the simulator, and scores the simulated surrogate after each simulated batch.
Its forecast is thus not a curve family but the outcome of actually executing the acquisition policy on the best available stand-in for the truth---connecting the forecasting half of the package back to the active-learning half.
The optimizer configuration and the oracle are supplied at construction; measure, search space, and pool are read from the task.

```{r simulate-forecaster}
sim_forecaster <- lrn("lce.simulate",
  optimizer = optimizer_al(
    learner = km_learner(),
    se_method = "auto",
    batch_size = 4L,
    n_candidates = 400L
  ),
  oracle_learner = km_learner(),
  n_restarts = 3L,
  seed = 1L
)
sim_forecaster$train(task_past)
sim_prediction <- sim_forecaster$predict_newdata(future_grid)

sim_table <- data.table(
  batch_nr = future_grid$batch_nr,
  simulated = signif(sim_prediction$response, 3)
)
merge(sim_table, realized[, .(batch_nr, realized = signif(mae, 3))],
  by = "batch_nr", all.x = TRUE)
```

The simulated curve reproduces the *shape* of the continuation but sits below the realized values.
This bias is inherent to the construction: the forecast scores the simulated surrogate against the oracle's own predictions, and the oracle---fit on 36 points---is smoother than the true response surface, so the simulated problem is easier than the real one.
The simulation forecast is therefore most valuable where the curve-fitting learners are blind by construction: it reacts to the acquisition policy, the batch size, and the candidate budget, and it produces joint sample paths (one per restart) whose spread reflects the stochasticity of the policy itself.

# Extrapolator Evaluation

Which forecaster should be trusted for a given learning curve?
Since extrapolators are learners, this is a resampling question, answered with the standard `mlr3` tools.
Two ingredients are specific to learning curves: the resampling scheme and the measures.

For resampling, random train--test splits would leak future batches into training.
`ResamplingLCE` (`rsmp("lce.expanding_cv")`) instead splits by whole batches in temporal order: each fold trains on all batches up to a moving cut-off and tests on the following `horizon` batches, with the cut-off advancing by `step_size` between folds.
The first cut-off, `min_train_batches`, has deliberately no default, since results are sensitive to how much history the forecasters see; it must also respect the learners' own minima (the parametric families need a handful of batches, `lce.spline_monotone` needs five).

The registered measures score per-batch forecasts:

```{r measure-zoo}
grep("^lce\\.", mlr_measures$keys(), value = TRUE)
```

`lce.mae`, `lce.mse`, and `lce.rmse` are point-forecast losses on the natural scale.
The distributional measures read the Gaussian-on-link predictive carried by the `"se"` predict type: `lce.crps` is the closed-form continuous ranked probability score (a proper score, our headline distributional criterion), `lce.coverage` and `lce.interval_score` assess central prediction intervals, `lce.pinball` scores predicted quantiles, and `lce.reach_brier` evaluates the predicted probability of having reached a target---the decision-oriented criterion matching [Section 8](#batches-to-target-forecasts-and-stopping-decisions).

We benchmark five forecasters on both learning-curve tasks, with folds forecasting two batches ahead:

```{r benchmark}
bench_learners <- list(
  lrn("lce.featureless"),
  lrn("lce.rolling_slope", window = 4L),
  lrn("lce.parametric_exponential"),
  lrn("lce.parametric_power_law"),
  lrn("lce.bootstrap", learner = lrn("lce.parametric_power_law"),
    n_bootstrap = 100L, seed = 1L)
)
for (learner in bench_learners) {
  learner$predict_type <- "se"
}

resampling <- rsmp("lce.expanding_cv",
  min_train_batches = 5L, horizon = 2L, step_size = 1L)

set.seed(5L)
benchmark_result <- benchmark(benchmark_grid(
  tasks = list(task_experiment, task_tree),
  learners = bench_learners,
  resamplings = resampling
))

aggregated <- benchmark_result$aggregate(
  list(msr("lce.mae"), msr("lce.crps"), msr("lce.coverage"))
)
aggregated[, .(task_id, learner_id,
  lce.mae = signif(lce.mae, 3),
  lce.crps = round(lce.crps, 3),
  lce.coverage = round(lce.coverage, 2))]
```

The two tasks reward different beliefs, as anticipated in [Section 5.4](#offline-curve-reconstruction).
On the steadily improving two-dimensional curve, the trend-following forecasters (rolling slope, power law) clearly beat the featureless baseline at both point and distributional accuracy.
On the plateaued tree curve the ranking inverts for point forecasts: predicting the best value seen so far is hard to beat once a curve has flattened, and the flexible extrapolators pay for occasionally chasing noise---although the distributional scores still favour the calibrated forecasters.
The practical reading is that no family dominates across traces, and that the forecaster feeding a costly stopping decision should first be validated on the trace types it will face; the benchmark above is precisely that validation, in miniature.
(With a single task and learner, `resample()` replaces `benchmark()` in the usual `mlr3` manner.)

# Batches to Target: Forecasts and Stopping Decisions

The question from the beginning of [Section 5](#learning-curve-construction)---*how many further batches?*---is answered by `lce_batches_to_target()`.
Given a trained forecaster, a future batch grid, and a performance target, it converts the per-batch predictive distributions into a distribution over the batch at which the target is first reached, reporting its quantiles, the full CDF over the grid, and the probability that the grid never reaches the target at all.
The optimization direction (here: lower is better) is read from the measure stored in the training task.

Two crossing semantics are available.
`crossing = "expected"` asks when the *mean curve* passes the target---the de-noised notion appropriate for "when will the model truly be good enough", computed from `se_epistemic`.
`crossing = "observed"` asks when the *realized, noisy* curve first dips below the target, the literal stopping time of a run that halts on its progress plot; being a first-passage property of a correlated sequence, it requires joint sample paths and hence a sample-based learner.
The observed crossing necessarily happens no later in distribution than the expected one, since noise can dip below the target early.

We set the target at a MAE of $0.2$ and forecast from batch 8, using the bootstrap power law of [Section 6.2](#distributional-forecasts) and, for contrast, a bootstrap around the rolling slope:

```{r batches-to-target}
mae_target <- 0.2

b2t_power_law <- lce_batches_to_target(forecaster,
  batch_grid = 9:24L, target = mae_target, crossing = "expected")
b2t_power_law_observed <- lce_batches_to_target(forecaster,
  batch_grid = 9:24L, target = mae_target, crossing = "observed")

slope_forecaster <- lrn("lce.bootstrap",
  learner = lrn("lce.rolling_slope"), n_bootstrap = 100L, seed = 1L)
slope_forecaster$train(task_past)
b2t_slope <- lce_batches_to_target(slope_forecaster,
  batch_grid = 9:24L, target = mae_target, crossing = "expected")

rbind(
  data.table(forecaster = "power law (expected)", t(b2t_power_law$quantiles),
    p_never = round(b2t_power_law$p_never, 3)),
  data.table(forecaster = "power law (observed)", t(b2t_power_law_observed$quantiles),
    p_never = round(b2t_power_law_observed$p_never, 3)),
  data.table(forecaster = "rolling slope (expected)", t(b2t_slope$quantiles),
    p_never = round(b2t_slope$p_never, 3))
)
```

In the run that actually continued, the target was first reached in batch `r if (run_code) realized[mae <= mae_target, min(batch_nr)] else "..."`.

```{r b2t-plot, fig.height = 4, fig.cap = "Figure 7: Batches-to-target forecasts made at batch 8 for a target MAE of 0.2: probability that the target has been reached, per batch, under two forecasters. The vertical line marks the batch in which the continued run actually reached the target."}
plot(b2t_power_law$grid$batch, b2t_power_law$grid$cdf, type = "s", lwd = 2,
  col = "darkorange3", ylim = c(0, 1), xlab = "batch",
  ylab = "P(target reached by batch)", main = "Batches-to-target forecast")
lines(b2t_slope$grid$batch, b2t_slope$grid$cdf, type = "s", lwd = 2,
  col = "steelblue4")
abline(v = realized[mae <= mae_target, min(batch_nr)], lty = 2, col = "grey40")
abline(h = c(0.1, 0.5, 0.9), lty = 3, col = "grey75")
legend("topleft", legend = c("bootstrap power law", "bootstrap rolling slope"),
  col = c("darkorange3", "steelblue4"), lwd = 2, bty = "n")
```

The two forecasters bracket the realized outcome: the rolling slope, extrapolating the recent trend, gives a median of about batch 12 with wide uncertainty and some probability mass on "not within the grid", while the more cautious power law places its median around batch 15, two to three batches after the realized crossing and with its lower decile just missing it.
This residual disagreement is the honest state of knowledge at batch 8, and the benchmark of [Section 7](#extrapolator-evaluation) is the tool for deciding which of the two to weight more heavily.
Either way the forecast is directly actionable: multiplying batches by the batch size of four converts the answer into simulator evaluations, which can be set against the cost of computing them and the value of reaching the target.

# Conclusion

The path taken by this vignette is the intended shape of a `celecx` analysis.
An objective wraps the expensive simulator ([Section 2](#a-first-active-learning-example)), an `OptimizerAL`---assembled by a factory or by hand---runs batch-sequential active learning on it ([Sections 2 to 4](#a-first-active-learning-example)), a callback or an offline replay condenses the run into a `TaskLCE` ([Section 5](#learning-curve-construction)), learning-curve learners fit and extrapolate that task with calibrated uncertainty ([Section 6](#learning-curve-extrapolation)), resampling and benchmarking select among them ([Section 7](#extrapolator-evaluation)), and `lce_batches_to_target()` turns the selected forecaster into a stopping decision ([Section 8](#batches-to-target-forecasts-and-stopping-decisions)).
Because every stage speaks the `mlr3` language, each of them can be swapped out or studied in isolation: active-learning strategies are optimizers, emulators and their uncertainty wrappers are regression learners, learning curves are tasks, and extrapolators are learners with measures and resampling schemes of their own.

Several parts of the package did not fit into this introduction.
Among the surrogates, wrappers for further Gaussian-process backends (`regr.gpfit`, `regr.hetgp`, `regr.tgp`, `regr.deepgp`) cover heteroscedastic and nonstationary emulation.
Among the pool tools, `ObjectivePoolRFun` and `ObjectivePoolWrapper` restrict live objectives to candidate lists, and the `ALDistance` dictionary (`clx_ald()`) underlying the distance-based acquisition functions extends them to mixed continuous--categorical spaces.
Among the forecasters, `lce.parametric_log`, `lce.parametric_logistic`, `lce.isotonic`, and `lce.spline_monotone` provide further curve shapes, the `"quantiles"` and `"target_reached"` predict types expose the predictive distribution in other formats, and the measures `lce.pinball`, `lce.interval_score`, and `lce.reach_brier` score them.
The help pages of the classes encountered above document these variations, and each dictionary (`mlr_space_samplers`, `mlr_al_distances`, and the `lce.` entries of `mlr_learners` and `mlr_measures`) can be listed and explored in the usual `mlr3` fashion.

# References {.unnumbered}

::: {#refs}
:::
