---
title: "Active Learning With Different Models in celecx"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1L)

library("R6")
library("mlr3")
library("mlr3learners")
library("data.table")
library("paradox")
library("bbotk")
library("celecx")

if (requireNamespace("lgr", quietly = TRUE)) {
  lgr::get_logger("bbotk")$set_threshold("warn")
  lgr::get_logger("mlr3")$set_threshold("warn")
  lgr::get_logger("mlr3mbo")$set_threshold("warn")
}
```

The main places where the learner is supplied are:

| Workflow | Where the model goes |
| --- | --- |
| Continuous uncertainty-based active learning | `optimize_active(..., learner = learner)` or `optimizer_active_learning(learner = learner)` |
| Pool methods GSy, iGS, QBC, IDEAL | `optimizer_pool_al(method, learner = learner)` |
| Pool methods GSx and random | no model is used |

If a learner supports standard errors natively, `se_method = "auto"` uses them.
If it only predicts responses, `se_method = "auto"` wraps it in
`LearnerRegrBootstrapSE`.

Note the pool methods GSy, iGS and IDEAL only need response predictions.

## Model factories

For repeated experiments, use small factory functions that return a fresh
learner. The active learning helpers clone the learner internally, but factories
make it harder to accidentally reuse a learner that was trained elsewhere.

```{r model_factories}
make_gp_learner <- function() {
  lrn("regr.km", covtype = "matern5_2", control = list(trace = FALSE))
}

make_rf_learner <- function() {
  lrn("regr.ranger", num.trees = 80L, min.node.size = 2L, num.threads = 1L)
}
```

The helper functions currently assert that the model inherits from
`LearnerRegr`. A plain `GraphLearner` from `mlr3pipelines` is a general
`Learner`, so this demo uses a tiny `LearnerRegr` wrapper for polynomial
regression. This is the same pattern to use for any custom model: implement
`private$.train()` and `private$.predict()`, then pass the object like any other
regression learner.

```{r polynomial_wrapper}
LearnerRegrPolynomialLM <- R6Class("LearnerRegrPolynomialLM",
  inherit = LearnerRegr,

  public = list(
    initialize = function(degree = 3L) {
      private$.degree <- checkmate::assert_int(degree, lower = 1L)
      super$initialize(
        id = sprintf("regr.poly_lm_d%i", private$.degree),
        param_set = ps(),
        predict_types = c("response", "se"),
        feature_types = c("integer", "numeric"),
        packages = "stats",
        label = sprintf("Polynomial linear model, degree %i", private$.degree),
        man = NA_character_
      )
    }
  ),

  private = list(
    .degree = NULL,

    .add_polynomial_terms = function(xdt) {
      xdt_poly <- data.table::copy(xdt)
      feature_ids <- names(xdt_poly)

      if (private$.degree >= 2L) {
        for (feature_id in feature_ids) {
          for (power in seq.int(2L, private$.degree)) {
            xdt_poly[, sprintf("%s_pow%i", feature_id, power) := get(feature_id)^power]
          }
        }
      }

      xdt_poly
    },

    .train = function(task) {
      data <- task$data(cols = c(task$target_names, task$feature_names))
      feature_data <- private$.add_polynomial_terms(
        data[, task$feature_names, with = FALSE]
      )
      model_data <- cbind(data[, task$target_names, with = FALSE], feature_data)

      stats::lm(
        stats::as.formula(sprintf("%s ~ .", task$target_names[[1L]])),
        data = model_data
      )
    },

    .predict = function(task) {
      feature_data <- private$.add_polynomial_terms(
        task$data(cols = task$feature_names)
      )
      prediction <- stats::predict(
        self$model,
        newdata = feature_data,
        se.fit = self$predict_type == "se"
      )

      if (self$predict_type == "se") {
        list(response = unname(prediction$fit), se = unname(prediction$se.fit))
      } else {
        list(response = unname(prediction))
      }
    }
  )
)

make_poly_learner <- function() {
  LearnerRegrPolynomialLM$new(degree = 3L)
}
```

Collect the model factories in one list so the rest of the demo can switch
models without changing the active learning code.

```{r learner_summary}
learner_factories <- list(
  "Gaussian process" = make_gp_learner,
  "Random forest" = make_rf_learner,
  "Polynomial regression" = make_poly_learner
)

learner_summary <- rbindlist(lapply(names(learner_factories), function(label) {
  learner <- learner_factories[[label]]()
  data.table(
    model = label,
    learner_id = learner$id,
    predict_types = paste(learner$predict_types, collapse = ", ")
  )
}))

learner_summary
```

## Continuous active learning

Here all three learners are plugged into the same continuous
uncertainty-based active learning workflow. The only changing argument is the
learner factory.

```{r continuous_setup}
objective_fun <- function(x) {
  sin(pi * x) + 0.15 * cos(3 * pi * x)
}

objective <- ObjectiveRFun$new(
  fun = function(xs) list(y = objective_fun(xs$x)),
  domain = ps(x = p_dbl(lower = 0, upper = 2)),
  codomain = ps(y = p_dbl(tags = "learn"))
)

continuous_grid <- data.table(x = seq(0, 2, length.out = 200L))
continuous_grid[, truth := objective_fun(x)]

make_continuous_optimizer <- function(learner) {
  optimizer <- optimizer_active_learning(
    learner = learner,
    se_method = "auto",
    batch_size = 1L,
    acq_evals = 40L
  )
  optimizer$param_set$set_values(n_init = 5L)
  optimizer
}
```

```{r continuous_run, message = FALSE, warning = FALSE, results = "hide"}
continuous_results <- lapply(learner_factories, function(factory) {
  optimize_active(
    objective = objective,
    n_evals = 14L,
    optimizer = make_continuous_optimizer(factory())
  )
})
```

```{r continuous_plot, fig.width = 8, fig.height = 9}
plot_continuous_run <- function(result, title) {
  archive_dt <- result$instance$archive$data
  pred_dt <- data.table::copy(continuous_grid)
  surrogate <- result$optimizer$surrogates$uncertainty
  prediction <- surrogate$predict(pred_dt[, .(x)])

  pred_dt[, mean := prediction$mean]
  pred_dt[, lower := mean - 1.96 * prediction$se]
  pred_dt[, upper := mean + 1.96 * prediction$se]

  plot(
    pred_dt$x,
    pred_dt$truth,
    type = "l",
    lwd = 2,
    col = "grey30",
    xlab = "x",
    ylab = "y",
    main = title
  )
  lines(pred_dt$x, pred_dt$mean, col = "navy", lwd = 2)
  lines(pred_dt$x, pred_dt$upper, col = "navy", lty = 2)
  lines(pred_dt$x, pred_dt$lower, col = "navy", lty = 2)
  points(archive_dt$x, archive_dt$y, pch = 19, col = "firebrick")
  text(archive_dt$x, archive_dt$y, labels = seq_len(nrow(archive_dt)), pos = 3, cex = 0.7)
}

old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))
invisible(Map(plot_continuous_run, continuous_results, names(continuous_results)))
```

For a response-only learner, request bootstrap standard errors explicitly:

```{r bootstrap_example, eval = FALSE}
optimizer_active_learning(
  learner = lrn("regr.rpart"),
  se_method = "bootstrap",
  n_bootstrap = 20L
)
```

## Pool active learning

Pool methods receive the learner when the optimizer is constructed. The
following example keeps the method fixed to iGS and changes only the learner.

```{r pool_setup}
pool_dt <- data.table(x = seq(0, 2, length.out = 61L))
pool_dt[, y := objective_fun(x)]

pool_objective <- ObjectiveDataset$new(
  dataset = pool_dt,
  domain = ps(x = p_dbl(lower = 0, upper = 2)),
  codomain = ps(y = p_dbl(tags = "learn"))
)

make_pool_optimizer <- function(learner) {
  optimizer_pool_al(
    "igs",
    learner = learner,
    n_init = 5L,
    batch_size = 1L
  )
}
```

```{r pool_run, message = FALSE, warning = FALSE, results = "hide"}
pool_results <- lapply(learner_factories, function(factory) {
  optimize_active(
    objective = pool_objective,
    n_evals = 14L,
    optimizer = make_pool_optimizer(factory())
  )
})
```

```{r pool_summary}
score_pool_result <- function(result, learner_factory) {
  eval_learner <- learner_factory()
  train_dt <- result$instance$archive$data[, .(x, y)]
  eval_task <- as_task_regr(train_dt, target = "y")
  eval_learner$train(eval_task)

  pred <- eval_learner$predict_newdata(pool_dt[, .(x)])$response
  sqrt(mean((pred - pool_dt$y)^2))
}

pool_summary <- rbindlist(Map(function(model_label, result, learner_factory) {
  data.table(
    model = model_label,
    method = "iGS",
    n_evals = result$instance$archive$n_evals,
    final_pool_rmse = score_pool_result(result, learner_factory)
  )
}, names(pool_results), pool_results, learner_factories))

pool_summary[order(final_pool_rmse)]
```

```{r pool_plot, fig.width = 8, fig.height = 9}
plot_pool_run <- function(result, title) {
  archive_dt <- result$instance$archive$data

  plot(
    pool_dt$x,
    pool_dt$y,
    type = "l",
    lwd = 2,
    col = "grey30",
    xlab = "x",
    ylab = "y",
    main = title
  )
  points(pool_dt$x, pool_dt$y, pch = 16, cex = 0.45, col = "grey70")
  points(archive_dt$x, archive_dt$y, pch = 19, col = "firebrick")
  text(archive_dt$x, archive_dt$y, labels = seq_len(nrow(archive_dt)), pos = 3, cex = 0.7)
}

old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))
invisible(Map(plot_pool_run, pool_results, names(pool_results)))
```

The same pattern applies to the other learner-based pool methods:

```{r other_pool_methods, eval = FALSE}
optimizer_pool_al("gsy", learner = make_rf_learner(), n_init = 5L)
optimizer_pool_al("ideal", learner = make_poly_learner(), n_init = 5L)
optimizer_pool_al("qbc", learner = make_rf_learner(), n_init = 5L, k_qbc = 5L)
```

GSx and random pool sampling are different because they do not use a model:

```{r model_free_pool_methods, eval = FALSE}
optimizer_pool_al("gsx", n_init = 5L)
optimizer_pool_al("random", n_init = 5L)
```
