---
title: "Active Learning Methods in celecx"
output: html_document
---

```{r prelim, eval = FALSE}
renv::restore()
devtools::install("..")

```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1L)

library("mlr3")
library("mlr3learners")
library("data.table")
library("bbotk")
library("celecx")

if (requireNamespace("lgr", quietly = TRUE)) {
  lgr::get_logger("bbotk")$set_threshold("warn")
  lgr::get_logger("mlr3")$set_threshold("warn")
  lgr::get_logger("mlr3mbo")$set_threshold("warn")
}

make_gp_learner <- function() {
  lrn("regr.km", covtype = "matern5_2", control = list(trace = FALSE))
}
```

This demo starts with a Gaussian process surrogate that explores an unknown one-dimensional function.
From there, it moves to a fixed candidate pool and compares the package's pool-based active learning methods: GSx, GSy, iGS (`optimizer_pool_al("igs")`), and IDEAL.

## Continuous active learning, as in the README

The default `optimize_active()` workflow is a good fit when the search space is continuous and the package should score a sampled set of candidate points with an uncertainty-based acquisition function.

```{r continuous_run, message = FALSE, warning = FALSE, results = "hide"}
objective <- ObjectiveRFun$new(
  fun = function(xs) list(y = sin(xs$x * pi)),
  domain = ps(x = p_dbl(lower = 0, upper = 2)),
  codomain = ps(y = p_dbl(tags = "learn"))
)

continuous_result <- optimize_active(
  objective = objective,
  n_evals = 10L,
  learner = make_gp_learner(),
  se_method = "auto",
  batch_size = 2L,
  acq_evals = 20L,
  multipoint_method = "greedy"
)

continuous_archive <- continuous_result$instance$archive$data
continuous_grid <- data.table(x = seq(0, 2, length.out = 200L))
continuous_grid[, truth := objective$fun(list(x = x))$y]

continuous_surrogate <- continuous_result$optimizer$surrogates$uncertainty
continuous_pred <- continuous_surrogate$predict(continuous_grid[, .(x)])
continuous_grid[, mean := continuous_pred$mean]
continuous_grid[, lower := mean - 1.96 * continuous_pred$se]
continuous_grid[, upper := mean + 1.96 * continuous_pred$se]
```

```{r continuous_plot, fig.width = 8, fig.height = 4.5}
batch_cols <- grDevices::hcl.colors(
  max(continuous_archive$batch_nr),
  palette = "Viridis"
)

plot(
  continuous_grid$x,
  continuous_grid$truth,
  type = "l",
  lwd = 2,
  col = "firebrick",
  xlab = "x",
  ylab = "y",
  main = "README-style active learning on sin(pi * x)"
)
lines(continuous_grid$x, continuous_grid$mean, col = "navy", lwd = 2)
lines(continuous_grid$x, continuous_grid$upper, col = "navy", lty = 2)
lines(continuous_grid$x, continuous_grid$lower, col = "navy", lty = 2)
points(
  continuous_archive$x,
  continuous_archive$y,
  pch = 19,
  cex = 1.1,
  col = batch_cols[continuous_archive$batch_nr]
)
text(
  x = continuous_archive$x,
  y = continuous_archive$y,
  labels = continuous_archive$batch_nr,
  pos = 3,
  cex = 0.8
)
legend(
  "bottomleft",
  legend = c("truth", "GP mean", "95% band", "queried points"),
  col = c("firebrick", "navy", "navy", "black"),
  lty = c(1, 1, 2, NA),
  pch = c(NA, NA, NA, 19),
  bty = "n"
)
```

The queried points are labeled by `batch_nr`, which makes it easy to see the initial design and the later uncertainty-driven batches.

## Switching to a fixed candidate pool

The following methods are most naturally used on fixed candidate pools.
They can also be applied to continuous objectives by scoring a random subsample via `pool_size`, but here we keep the fixed-pool setting and run them sequentially.

```{r pool_setup}
pool_dt <- data.table(x = seq(0, 2, length.out = 61L))
pool_dt[, y := sin(pi * x) + 0.15 * cos(3 * pi * x)]

pool_objective <- ObjectiveDataset$new(
  dataset = pool_dt,
  domain = ps(x = p_dbl(lower = 0, upper = 2)),
  codomain = ps(y = p_dbl(tags = "learn"))
)

plot(
  pool_dt$x,
  pool_dt$y,
  type = "l",
  lwd = 2,
  col = "grey30",
  xlab = "x",
  ylab = "y",
  main = "Fixed candidate pool used for the method comparison"
)
points(pool_dt$x, pool_dt$y, pch = 16, cex = 0.45, col = "grey65")
```

The four methods differ in how they score the next candidate:

| Method | Constructor | Main idea | Learner required |
| --- | --- | --- | --- |
| GSx | `optimizer_pool_al("gsx")` | spread points out in input space | no |
| GSy | `optimizer_pool_al("gsy", learner = ...)` | prefer predicted outputs far from queried responses | yes |
| iGS | `optimizer_pool_al("igs", learner = ...)` | combine input-space and output-space separation | yes |
| IDEAL | `optimizer_pool_al("ideal", learner = ...)` | combine model residuals with distance-based exploration | yes |


```{r pool_methods, message = FALSE, warning = FALSE, results = "hide"}
pool_methods <- list(
  GSx = optimizer_pool_al("gsx", n_init = 4L, batch_size = 1L),
  GSy = optimizer_pool_al("gsy", learner = make_gp_learner(), n_init = 4L),
  iGS = optimizer_pool_al("igs", learner = make_gp_learner(), n_init = 4L),
  IDEAL = optimizer_pool_al(
    "ideal",
    learner = make_gp_learner(),
    n_init = 4L,
    delta = 1,
    batch_size = 1L
  )
)

pool_results <- lapply(pool_methods, function(optimizer) {
  optimize_active(
    objective = pool_objective,
    n_evals = 12L,
    optimizer = optimizer
  )
})

score_pool_result <- function(result, pool_dt) {
  eval_learner <- make_gp_learner()
  train_dt <- result$instance$archive$data[, .(x, y)]
  eval_task <- as_task_regr(train_dt, target = "y")
  eval_learner$train(eval_task)

  pred <- eval_learner$predict_newdata(pool_dt[, .(x)])$response
  sqrt(mean((pred - pool_dt$y)^2))
}

pool_summary <- rbindlist(Map(function(method_label, result) {
  data.table(
    method = method_label,
    n_evals = result$instance$archive$n_evals,
    final_rmse = score_pool_result(result, pool_dt)
  )
}, names(pool_results), pool_results))
```

```{r pool_summary}
pool_summary[order(final_rmse)]
```

```{r pool_plots, fig.width = 10, fig.height = 8}
plot_pool_run <- function(result, title) {
  archive_dt <- result$instance$archive$data
  point_cols <- grDevices::hcl.colors(nrow(archive_dt), palette = "Viridis")

  plot(
    pool_dt$x,
    pool_dt$y,
    type = "l",
    lwd = 2,
    col = "grey30",
    xlab = "x",
    ylab = "y",
    main = title
  )
  points(
    archive_dt$x,
    archive_dt$y,
    pch = 19,
    cex = 1.1,
    col = point_cols
  )
  text(
    x = archive_dt$x,
    y = archive_dt$y,
    labels = seq_len(nrow(archive_dt)),
    pos = 3,
    cex = 0.75
  )
}

old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
invisible(Map(plot_pool_run, pool_results, names(pool_results)))
```

The labels mark evaluation order, so the different selection strategies are visible directly in the plots:

- GSx pushes coverage in input space first.
- GSy chases places where the surrogate predicts outputs far from what has   already been observed.
- iGS tries to keep both input-space and output-space diversity.
- IDEAL mixes geometry with a model-based uncertainty estimate derived from inverse-distance-weighted residuals.
