---
title: "Tracking Active Learning Performance in celecx"
output: html_document
---

```{r prelim, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install("..", dependencies = TRUE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1L)

library("mlr3")
library("mlr3learners")
library("mlr3misc")
library("data.table")
library("bbotk")
library("celecx")

if (requireNamespace("lgr", quietly = TRUE)) {
  lgr::get_logger("bbotk")$set_threshold("warn")
  lgr::get_logger("mlr3")$set_threshold("warn")
  lgr::get_logger("mlr3mbo")$set_threshold("warn")
}
```

This demo attaches a surrogate performance callback to an active learning run.
After every evaluated batch, including the initial design, the callback evaluates
the named surrogate on a held-out regression task and stores the scores in a
`data.table`.

## Objective and test task

```{r objective}
objective_fun <- function(x) {
  sin(pi * x) + 0.15 * cos(3 * pi * x)
}

objective <- ObjectiveRFun$new(
  fun = function(xs) list(y = objective_fun(xs$x)),
  domain = ps(x = p_dbl(lower = 0, upper = 2)),
  codomain = ps(y = p_dbl(tags = "learn"))
)

test_dt <- data.table(x = seq(0, 2, length.out = 200L))
test_dt[, y := objective_fun(x)]
test_task <- as_task_regr(test_dt, target = "y")
```

## Run with performance tracking

```{r run, message = FALSE, warning = FALSE, results = "hide"}
perf <- clbk("celecx.surrogate_performance",
  surrogate_id = "uncertainty",
  task = test_task,
  measures = list(
    r2 = msr("regr.rsq"),
    mae = msr("regr.mae"),
    rmse = msr("regr.rmse")
  )
)

result <- optimize_active(
  objective = objective,
  n_evals = 20L,
  learner = lrn("regr.rpart"),
  n_bootstrap = 8L,
  batch_size = 2L,
  acq_evals = 40L,
  callbacks = list(perf)
)

performance_dt <- perf$data
```

```{r data}
performance_dt
```

## Plot progress

```{r plot, fig.width = 9, fig.height = 4}
old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(
  performance_dt$n_evals,
  performance_dt$r2,
  type = "b",
  pch = 19,
  col = "navy",
  xlab = "Evaluations",
  ylab = "R-squared",
  main = "Held-out R-squared"
)
abline(h = 1, lty = 2, col = "grey70")

matplot(
  performance_dt$n_evals,
  performance_dt[, .(mae, rmse)],
  type = "b",
  pch = c(19, 17),
  lty = 1,
  col = c("firebrick", "darkgreen"),
  xlab = "Evaluations",
  ylab = "Error",
  main = "Held-out error"
)
legend(
  "topright",
  legend = c("MAE", "RMSE"),
  col = c("firebrick", "darkgreen"),
  pch = c(19, 17),
  lty = 1,
  bty = "n"
)
```

The performance table remains available from the callback after optimization:

```{r readout}
perf$latest
```
